---
title: "Case Study Brain Regions"
author: "Chenghao Zhu"
date: "`r strftime(Sys.time(), format = '%Y-%m-%d')`"
output: 
    html_document:
        code_folding: show
        toc: true
        toc_float: true
runtime: "shiny"
---

<style type="text/css">
code {
  font-family: monaco;
}
body p{
    font-size: 12pt
}
body li {
    font: 12pt
}
.datatables{
    overflow: auto;
    white-space: nowrap;
}
</style>

<hr>

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, message=FALSE, warning=FALSE}
options(repos = BiocManager::repositories())
library(tidyverse)
library(reshape2)
library(HTSet)
library(ggplot2)
library(plotly)
library(DT)
library(limma)
library(edgeR)
library(shiny)
```


```{r}
# loading data
abs_abund0 = readRDS("data/case_study2.rds")$abs_raw
abs_abund_mgs <- readRDS("data/case_study2.rds")$abs_mgs

# set default ggplot theme
theme_set(theme_bw())
# set stringsAsFactors as FALSE by default
options(stringsAsFactors = FALSE)
```

```{r, echo = FALSE}
# tags$div(
#     class = "row",
#     tags$div(
#         class="col-sm-6",
#         selectInput("input-data-type", label = "Data Type", choices = c("raw", "mgtissues"))
#     )
# )
# abs_abund0 <- reactive({
#   if(input$`input-data-type` == "raw"){abs_abund_raw}else{abs_abund_mgs}
# })
```

<hr>

## 1. Data clarification

The excel sheet received `CaseStudy_AD_20191116_R&A.xlsx` has three sheets named `MTC_AbsAbund`, `LPFc_AbsAbund`, and `LPbC_AbsAbund`, that each has the data for each of the three brain regions analyzed. It seems that different glycans were monitored in the three brain regions, so the rows of the three sheets do not match. The `LPFc_AbsAbund` sheet has 240 rows, the `MTC_AbsAbund` has 258 rows, and the `LPbC_AbsAbund` has 265 rows. I read all three sheets in, and merged them using the `Glycan` column. The merged dataset has 293 glycans (rows). The glycans not observed in a region but present in another region are set to zero.

## 2. Zeros/missing values

This dataset also contains some zeros/missing values.

```{r}
  (data.frame(
    feature = featureNames(abs_abund0),
    abs_abund = apply(abs_abund0$edata, 1, function(row){sum(row == 0)})
) %>% 
    ggplot() + 
    geom_histogram(aes(x = abs_abund), color = "white", bins = 25, fill = "steelblue") +
    labs(x = "number of zeros")) %>%
    ggplotly

```

For each brain regions, we filtered glycans by reponse intensity: Glycans with > 0 intensity in 50% of samples and in at leat 3 samples from each group were used for further analysis.  
Relative abundance were also calculated with the sum of selected glycans as the denominator in each brain region.

```{r}
min.obs.freq = 0.50
# filter_by_missing = function(mset, region, cutoff){
#     samples = which(mset$pdata$region == region)
#     mset = subset_samples(mset, sampleNames(mset)[samples])
#     mset = subset_features(mset, apply(mset$edata, 1, function(x) 
#         sum(x != 0) > cutoff * length(samples)
#     ))
#     mset
# }
filter_by_missing = function(mset, region, cutoff){
    samples_con <- which(mset$pdata$region == region & mset$pdata$group == "Control")
    mset_con = subset_samples(mset, sampleNames(mset)[samples_con])
    mset_con = subset_features(mset_con, apply(mset_con$edata, 1, function(x) 
        sum(x != 0) > 3
    ))
    samples_AD <- which(mset$pdata$region == region & mset$pdata$group == "AD")
    mset_AD = subset_samples(mset, sampleNames(mset)[samples_AD])
    mset_AD = subset_features(mset_AD, apply(mset_AD$edata, 1, function(x) 
        sum(x != 0) > 3
    ))
    features <- intersect(featureNames(mset_con), featureNames(mset_AD))
    samples = which(mset$pdata$region == region)
    mset <- subset_samples(mset, sampleNames(mset)[samples])
    mset <- subset_features(mset, apply(mset$edata, 1, function(x)
      sum(x != 0) > cutoff * length(samples)
    ))
    mset <- subset_features(mset, featureNames(mset) %in% features)
    mset
}
regions = unique(abs_abund0$pdata$region) 
abs_abund = lapply(regions, function(region) filter_by_missing(abs_abund0, region, min.obs.freq))
names(abs_abund) = regions
# regions = unique(abs_abund0$pdata$region) 
# abs_abund = lapply(regions, function(region) filter_by_missing(abs_abund0, region, min.obs.freq))
# names(abs_abund) = regions
rel_abund = lapply(abs_abund, function(mset) 
  transform_by_sample(mset, function(x) (x + 1)/sum(x+1)))
names(rel_abund) = names(abs_abund)

```

## 3. Normality

As shown below in the histogram of the z score scaled median of each glycan, it skews to the left due to the zeros. The z score scaled median was used to pick a representative point from each glycan. These histograms helped to check if glycan data satisfied the linear model assumption (normal distribution).

```{r, echo = FALSE}
tags$div(
    class = "row",
    tags$div(
        class="col-sm-6",
        selectInput("data-type_hist", label = "Data Type", choices = c("abs_abund", "rel_abund"))
    ),
    tags$div(
        class="col-sm-6",
        selectInput("region_hist", label = "Brain Region", choices = c("LCBC", "MTC", "LPFC"))
    )
)
```


```{r}
output$hist <- renderPlotly({
  if (input$`data-type_hist` == "abs_abund") {
    mset <- abs_abund
  } else {
    mset <- rel_abund
  }
  (data.frame(
    glycan = featureNames(mset[[input$region_hist]]),
    abs_abund = apply(mset[[input$region_hist]]$edata, 1, function(x) median(scale(x)))
  ) %>% 
    ggplot(aes(x = abs_abund, y = ..density..)) +
    geom_histogram(color = "white", bins = 25, fill = "steelblue") +
    geom_density()+
    labs(x = input$`data-type-hist`)) %>%
    ggplotly
})
plotlyOutput("hist")
```

Again, taking a log transformation makes it look better for absolute abundance, but not for relative abundance.

```{r}
output$hist_log <- renderPlotly({
  if (input$`data-type_hist` == "abs_abund") {
    mset <- abs_abund
  } else {
    mset <- rel_abund
  }
  (data.frame(
    glycan = featureNames(mset[[input$region_hist]]),
    abs_abund = apply(mset[[input$region_hist]]$edata, 1, function(x) median(scale(log(x+1))))
  ) %>% 
    ggplot(aes(x = abs_abund, y = ..density..)) +
    geom_histogram(color = "white", bins = 25, fill = "steelblue") +
    geom_density()+
    labs(x = input$`data-type-hist`)) %>%
    ggplotly
})

# (data.frame(
#     glycan = featureNames(abs_abund0),
#     abs_abund = apply(abs_abund0$conc_table, 1, function(x) median(scale(log(x + 1))))
# ) %>% 
#     ggplot(aes(x = abs_abund, y = ..density..)) +
#     geom_histogram(color = "white", bins = 25, fill = "steelblue") +
#     geom_density()) %>%
#     ggplotly
plotlyOutput("hist_log")
```


<hr>

## 4. Princicple component analysis

```{r, eval = FALSE, include = FALSE}
pca1 = abs_abund %>% conc_table %>% log %>%  apply(1, scale) %>% prcomp
pca2 = rel_abund  %>% conc_table %>% log %>%  apply(1, scale) %>% prcomp
df = data.frame(pca1$x[,1:2], pca2$x[,1:2]) %>%
    `colnames<-`(c(
        paste0("abs_abund-", c("PC1", "PC2")),
        paste0("rel_abund-", c("PC1", "PC2"))
    )) %>%
    mutate(
        region = abs_abund$sample_table$region,
        sample = sampleNames(abs_abund),
        group = abs_abund$sample_table$group
    ) %>%
    melt(id.vars = c("sample", "region", "group")) %>%
    separate(variable, into = c("data type", "PC"), sep = "-") %>%
    dcast(sample + region + `data type` + group ~ PC)
```

```{r, out.width="100%", eval = FALSE, include = FALSE}
(ggplot(df) +
    geom_point(aes(x = PC1, y = PC2, color = group)) +
    facet_wrap(~`data type`, scales = "free")) %>%
    ggplotly
```

```{r, out.width="100%", eval = FALSE, include = FALSE}
(ggplot(df) +
    geom_point(aes(x = PC1, y = PC2, color = region)) +
    facet_wrap(~`data type`, scales = "free")) %>%
    ggplotly
```

```{r, eval = FALSE, include = FALSE, echo = FALSE}
subset_pca = function(mset, region){
    mset = subset_samples(mset, mset$sample_table$region == region)
    mset = subset_features(mset, apply(mset$conc_table, 1, function(x) sum(x > 1) > 0))
    pca = mset %>% conc_table %>% log %>%  apply(1, scale) %>% prcomp
    p = pca$x[,1:2] %>% as.data.frame %>% 
        mutate(group = mset$sample_table$group) %>%
        ggplot(aes(PC1, PC2, color = group)) +
        geom_point()
    ggplotly(p)
}
subset_pca(rel_abund, "MTC")
```

## 5. Linear Model

Fit the linear model for each brain region with log-transformed absolute abundance and relative abundance.

```{r}
fill_na = function(mset){
    transform_by_feature(mset, function(x){
        x[x == 0] = min(x[x != 0])/2
        return(x)
    })
}
lm = list()
# abs_abund
regions = names(abs_abund)
lm$abs_abund = lapply(regions, function(reg){
    mset = fill_na(abs_abund[[reg]])
    edata = log(mset$edata)
    pdata = mset$pdata
    pdata$group = factor(pdata$group)
    pdata$group = relevel(pdata$group, "Control")
    design = model.matrix(~group, data = pdata)
    fit = lmFit(edata, design)
    fit = eBayes(fit)
    topTable(fit, coef = "groupAD", number = Inf, sort.by = "none")
})
names(lm$abs_abund) = regions

# rel_abund
lm$rel_abund = lapply(regions, function(reg){
    mset = rel_abund[[reg]]
    edata = log(mset$edata)
    pdata = mset$pdata
    pdata$group = factor(pdata$group)
    pdata$group = relevel(pdata$group, "Control")
    design = model.matrix(~group, data = pdata)
    fit = lmFit(edata, design)
    fit = eBayes(fit)
    topTable(fit, coef = "groupAD", number = Inf, sort.by = "none")
})
names(lm$rel_abund) = regions


```

```{r, echo = FALSE}
tags$div(
    class = "row",
    tags$div(
        class="col-sm-6",
        selectInput("data-type", label = "Data Type", choices = names(lm))
    ),
    tags$div(
        class="col-sm-6",
        selectInput("region", label = "Brain Region", choices = c("LCBC", "MTC", "LPFC"))
    )
)
```

```{r, echo = FALSE}
dataTableOutput("dt")
output$dt = renderDataTable({
    if(input$`data-type` == "abs_abund"){
        mset = abs_abund[[input$region]]
    } else {
        mset = abs_abund[[input$region]]
    }
    tt = lm[[input$`data-type`]][[input$region]]
    datatable(
        cbind(mset$fdata, tt),
        options = list(order = list(11, "asc")),
        selection = list(mode = "single", selected = 1)
    ) %>%
        formatSignif(
            columns = c("logFC", "AveExpr", "t", "P.Value", "adj.P.Val", "B"),
            digits = 4
        ) 
})
```

```{r, echo = FALSE}
downloadLink("downloadData", "Download Excel", class="btn btn-info btn-xs disabled")
output$downloadData = downloadHandler(
    'case_study_lm_results.xlsx.xlsx',
    content = function(file) {
        file.copy('case_study_lm_results.xlsx', file)
    },
    contentType = "application/zip"
)
```

<a class="btn btn-danger btn-xs" role="button" data-toggle="collapse" href="#excel" aria-expanded="false" aria-controls="excel" style="float: right;">
export to excel
</a>
<div class="collapse" id="excel">
```{r, eval = F, include = FALSE}
Sys.setenv(JAVA_HOME='/Library/Java/JavaVirtualMachines/jdk-11.0.1.jdk/Contents/Home')
library(xlsx)
wb = createWorkbook(type='xlsx')
style_header = CellStyle(wb) + Font(wb, isBold = TRUE)
border = function(position){
    return(Border(color = "black", position = position, pen = "BORDER_THIN"))
}
for(type in names(lm)){
    sheet = createSheet(wb, sheetName = type)
    setColumnWidth(sheet, colIndex = 1, colWidth = 20)
    title_row = createRow(sheet, 1)
    for(i in seq_along(lm[[type]])){
        colIndex = (i - 1) * 6 + 2
        cell = createCell(title_row, colIndex)
        addMergedRegion(sheet, 1,1,colIndex, colIndex + 5)
        setCellValue(cell[[1]], names(lm[[type]])[i])
        addDataFrame(lm[[type]][[i]], sheet, startRow = 2,
                     startColumn = colIndex - (i == 1), row.names = i == 1)
    }
    
    rows = getRows(sheet)
    cells = getCells(rows)
    nrow = nrow(lm[[type]]$MTC) + 1
    ncol = 6 * length(lm[[type]]) + 1
    
    for(k in seq_along(cells)){
        style = CellStyle(wb)
        if(k <= length(lm[[type]])){
            style = style_header + border(c("TOP", "BOTTOM", "RIGHT", "LEFT"))
        } else {
            i = as.integer((k - length(lm[[type]]) - 1) / ncol) + 1
            #j = (k - length(lm[[type]])) %% ncol
            j = (k - length(lm[[type]])) - (i - 1) * ncol
            if(i == 1){
                if(j == 1){
                    style = style_header + border(c("TOP", "BOTTOM", "RIGHT", "LEFT"))
                } else if(j %% 6 == 1) {
                    style = style_header + border(c("TOP", "BOTTOM", "RIGHT"))
                } else{
                    style = style_header + border(c("TOP", "BOTTOM"))
                }
            } else if (i != nrow) {
                if(j == 1){
                    style = style_header + border(c("RIGHT", "LEFT"))
                } else if(j %% 6 == 1) {
                    style = style_header + border("RIGHT")
                } else {
                    next
                }
            } else {
                if(j == 1){
                    style = style_header + border(c("BOTTOM", "RIGHT", "LEFT"))
                } else if(j %% 6 == 1) {
                    style = style + border(c("BOTTOM", "RIGHT"))
                } else {
                    style = style + border("BOTTOM")
                }
            }
        }
        setCellStyle(cells[[k]], style)
    }
}
saveWorkbook(wb, 'case_study_lm_results.xlsx')
```
</div>

## 6. Visualization {.tabset}

<div class="alert alert-info">Click on the tabs to see differet plots. The glycan in the boxplox is linked to the table above. Click on the row to see the effect.</div>

### Volcano plot

```{r echo = FALSE}
renderPlotly({
    lm[[input$`data-type`]][[input$region]] %>%
        ggplot() + 
        geom_point(aes(x = logFC, y = -log(P.Value)), color = "grey10", alpha = 0.6) +
        geom_hline(yintercept = -log(0.05), color = "red", linetype = "dashed")
})
```

### Histogram

```{r echo = FALSE}
renderPlotly({
    lm[[input$`data-type`]][[input$region]] %>%
        ggplot() +
        geom_histogram(aes(x = P.Value), bins = 40, color = "white", fill = "grey25") +
        geom_vline(xintercept = 0.05, color = "red", linetype = "dashed")
})
```

### Box plot

```{r echo = FALSE}
renderPlotly({
    if(input$`data-type` == "abs_abund"){
        mset = abs_abund[[input$region]]
    } else {
        mset = abs_abund[[input$region]]
    }
    title = featureNames(mset)[input$dt_rows_selected]
    data.frame(
        value = mset$edata[input$dt_rows_selected,],
        group = mset$pdata$group,
        region = input$region
    ) %>%
        filter(region == input$region) %>%
        ggplot(aes(group, value)) +
        geom_boxplot(aes(fill = group)) +
        geom_point(shape = 21, fill = "grey", color = "black", size = 2) +
        labs(title = title, x = "", y = "")
})
```


## 6. Conclustion

There doesn't seem to be a majore separation between either patient groups or brain regions according to the PCA plot. However, the linear regression result showed that, the relative abundance of 190 glycans in MTC were significantly different in patient groups, most of them are higher in the AD patiens. 

## 7. Session Information

```{r}
sessionInfo()
```